Parsing Postfix log files

### 2007-11-06 doc/logparser.tex
Replaced SMTP with \SMTP{} everywhere. Put the included tables into table environments and added a list of tables. Added the glossary and made it and the acronym list play together.

\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
% Redefine the command used to produce the glossary title, because the
% default command produces an unnumbered section whereas I want a numbered
% section.
\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
\makeglossary{}
\makeacronym{}

@@

@@

\newpage
\tableofcontents
\listoffigures

@@

The bibliography contains references to the resources used in developing
the algorithm, writing the program and preparing this paper.  Also listed
are some additional resources expected to be helpful in understanding SMTP,
Postfix, anti-spam techniques, or the paper.

@@

The \SMTPlong, originally defined in \RFC{} 821~\cite{RFC821} and updated
in \RFC{} 2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{} 760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time.}  A typical SMTP conversation
resembles the following (the lines starting with a three digit number are
sent by the server, all other lines are sent by the client):

@@

A detailed description of SMTP is beyond the scope of this document:
introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table~\cite{cidr_table}}
will check whether the IP address of the connected client matches the left
hand side of each line in the file and will return the right hand side of
the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the IP address of the client is present in the listed RBL\@.

@@

A policy server~\cite{policy-servers} is an external program that accepts
state information from Postfix for each SMTP command not rejected by an
earlier restriction; the policy server can utilise that state information
to implement whatever logic is required.  E.g.\ some users can be
restricted to sending mail on the third Tuesday after pay day only --- this
example may actually be useful in a payroll system to prevent problems from
spam or (worse) phishing mails with faked sender addresses.  More commonly
encountered scenarios are:

@@

    \item Checking Sender Policy Framework records~\cite{openspf,
        wikipedia-spf}.  SPF records specify which mail servers are allowed
        to send mail claiming to be from a particular domain.  The
        intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe-jobs~\cite{wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of SMTP\@.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on the second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        retrying after a temporary failure is uneconomical for a spammer,
        but that a legitimate mail server must retry.  Sadly spammers are
        using increasingly complex and well written programs to distribute
        spam, frequently using an ISP provided SMTP server from a
        compromised machine on the ISP's network.  Greylisting is slowly
        becoming less useful, but it does block a large percentage of spam
        mail at the moment.

@@

The data stored is limited to time and date, hostname and IP address of
client, sender and recipient addresses, and the SMTP code.  Handling of
multiple recipients, SMTP codes or remote servers\footnote{A single mail
may be sent to multiple remote servers if it was addressed to recipients in
different domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained in the paper, but the options appear to be:

@@

    \item The SMTP protocol.

@@

    \item [helo] The hostname used in the HELO command.  The HELO
        occasionally changes during a connection, presumably because spam
        or virus senders think it's a good idea.  By default Postfix only
        logs the HELO when it rejects an SMTP command, but it is quite easy
        to rectify this:

@@

Every log line corresponding to Postfix performing an action other than
INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
SMTP command, delivering a mail, or bouncing a mail.  Any log line where
the Postfix action is INFO is not interesting in and of itself, but
provides additional information which will be saved with other results;
IGNORED log lines don't even provide useful information.  Each row is
associated with a single connection, though there may be many results per
connection.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an SMTP command --- a
        facility that is quite useful for testing new restrictions.  This
        field will be 1 if the log line parsed was a warning rather than a
        real rejection, or 0 for a real rejection or any other result.

    \item [smtp\_code] The SMTP code associated with the log line.  In
        general an SMTP code is only present for a rejection or final
        delivery; results missing an SMTP code will duplicate the SMTP code
        of other results in the connection.  Some final delivery log lines
        don't contain an SMTP code: in those cases the code is faked based
        on the success or failure represented by the log line.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's regex, generate an NFA, and compare all NFAs
to determine if any overlap.  The author has not attempted to do this: such
a project would probably qualify for a PhD, and may involve solving the
Halting Problem and circumventing the Church-Turing Thesis.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line determines which rule
matches.  Overlapping rules are frequently a requirement, allowing a more
specific rule to match some lines and a more general rule to match the
majority, e.g.\ separating smtp delivery to a filtering proxy from smtp
delivery to the rest of the world.  The algorithm provides a facility for
ordering overlapping rules: the priority field in each rule (defaults to
zero).  Rules are tried in order of priority, highest first, overriding the
hits field maintained by the parser.  Negative priorities may also be
useful for catchall rules.

@@

    \item Mail enters the system via SMTP or local submission.

@@

Everything starts off with a mail entering the system, whether by local
submission via postdrop/sendmail, by SMTP, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: pickup; flowchart:~2).

SMTP is more complicated:

@@

    \item The client disconnects (action: disconnect; flowchart:~6).  If
        Postfix has rejected any SMTP commands the data will be saved to
        the database; if not there won't be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

@@

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: track; flowchart:~11).\footnote{Relayed
        delivery is performed by the SMTP client; local delivery means
        local to the server, i.e.\ an address the server is final
        destination for.} In this case the mail may already have been
        created (action: qmgr\_chooses\_mail; flowchart:~3; described
        above) and the unknown origin flag will be cleared; if not a new
        data structure will be created.  In both cases the new mail is
        marked as a child of the parent.  The log line in question is:

@@

    \item [REJECTION] Deal with postfix rejecting an SMTP command from the
        remote client: log the rejection with a mail if there is a queueid
        in the log line, or with the connection if not.

@@

    \item Neither smtpd nor pickup have logged any messages associated with
        the mail, indicating it was generated internally by Postfix, not
        accepted via SMTP or submitted locally by sendmail.

@@

Some mail clients appear to send the following sequence of commands during
the SMTP session:

@@

The DATA phase of the SMTP conversation is where the headers and body of the
mail are transferred.  Sometimes there is a timeout or the connection is
lost\footnote{For brevity's sake timeout will be used throughout this
section, but everything applies equally to lost connections.} during the
DATA phase; when this occurs Postfix will discard the mail and the parser
needs to discard the data associated with that mail.  It seems more
intuitive to save the data to the database, but if a timeout occurs there
won't yet have been any data gathered for the mail, so there is none
available to save; the timeout is saved with the connection data instead.

@@

            \item For mail sent: to addresses, SMTP code, and delay.

@@

\input{build/stats-normal-filesize-line-count-include}

@@

The difference between logs with a mail loop and logs without can be seen
in this table showing the increases for the different rule orderings and
combinations of log files:

\noindent\input{build/stats-normal-shuffle-reverse-include.tex}

@@

\showgraph{build/plot-cached-discarded-factor.ps}{Regex: percentage
execution time increase}{normal regex vs discarded regex factor}

@@

\input{build/stats-cached-discarded-include-for-graph}

@@

% The appropriate command has been redefined so this will produce a section
% and label called glossary.
\printglossary{}
\printacronym{}

>>>>

\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
\makeglossary{}
\makeacronym{}

@@

%\showtable{filename}{caption}{label}
\newcommand{\showtable}[3]{
    \begin{table}[ht]
        \caption{#2}\label{#3}
        \input{#1}
    \end{table}
}

@@

\newpage
\tableofcontents
\listoffigures
\listoftables

@@

The bibliography contains references to the resources used in developing
the algorithm, writing the program and preparing this paper.  Also listed
are some additional resources expected to be helpful in understanding
\SMTP{}, Postfix, anti-spam techniques, or the paper.

@@

The \SMTPlong, originally defined in \RFC{} 821~\cite{RFC821} and updated
in \RFC{} 2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{} 760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time.}  A typical \SMTP{}
conversation resembles the following (the lines starting with a three digit
number are sent by the server, all other lines are sent by the client):

@@

A detailed description of \SMTP{} is beyond the scope of this document:
introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table~\cite{cidr_table}}
will check whether the IP address of the connected client matches the left
hand side of each line in the file and will return the right hand side of
the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the IP address of the client is present in the listed RBL\@.

@@

A policy server~\cite{policy-servers} is an external program that accepts
state information from Postfix for each \SMTP{} command not rejected by an
earlier restriction; the policy server can utilise that state information
to implement whatever logic is required.  E.g.\ some users can be
restricted to sending mail on the third Tuesday after pay day only --- this
example may actually be useful in a payroll system to prevent problems from
spam or (worse) phishing mails with faked sender addresses.  More commonly
encountered scenarios are:

@@

    \item Checking Sender Policy Framework records~\cite{openspf,
        wikipedia-spf}.  SPF records specify which mail servers are allowed
        to send mail claiming to be from a particular domain.  The
        intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe-jobs~\cite{wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}\@.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on the second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        retrying after a temporary failure is uneconomical for a spammer,
        but that a legitimate mail server must retry.  Sadly spammers are
        using increasingly complex and well written programs to distribute
        spam, frequently using an ISP provided \SMTP{} server from a
        compromised machine on the ISP's network.  Greylisting is slowly
        becoming less useful, but it does block a large percentage of spam
        mail at the moment.

@@

The data stored is limited to time and date, hostname and IP address of
client, sender and recipient addresses, and the \SMTP{} code.  Handling of
multiple recipients, \SMTP{} codes or remote servers\footnote{A single mail
may be sent to multiple remote servers if it was addressed to recipients in
different domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained in the paper, but the options appear to be:

@@

    \item The \SMTP{} protocol.

@@

    \item [helo] The hostname used in the HELO command.  The HELO
        occasionally changes during a connection, presumably because spam
        or virus senders think it's a good idea.  By default Postfix only
        logs the HELO when it rejects an \SMTP{} command, but it is quite
        easy to rectify this:

@@

Every log line corresponding to Postfix performing an action other than
INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
\SMTP{} command, delivering a mail, or bouncing a mail.  Any log line where
the Postfix action is INFO is not interesting in and of itself, but
provides additional information which will be saved with other results;
IGNORED log lines don't even provide useful information.  Each row is
associated with a single connection, though there may be many results per
connection.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \SMTP{} command --- a
        facility that is quite useful for testing new restrictions.  This
        field will be 1 if the log line parsed was a warning rather than a
        real rejection, or 0 for a real rejection or any other result.

    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
        general an \SMTP{} code is only present for a rejection or final
        delivery; results missing an \SMTP{} code will duplicate the
        \SMTP{} code of other results in the connection.  Some final
        delivery log lines don't contain an \SMTP{} code: in those cases
        the code is faked based on the success or failure represented by
        the log line.

@@

The parser does not try to detect overlapping rules;\footnote{It may be
possible to parse each rule's regex, generate an NFA, and compare all NFAs
to determine if any overlap.  The author has not attempted to do this: such
a project would probably qualify for a PhD, and may involve solving the
Halting Problem and circumventing the Church-Turing Thesis.} that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the order in which rules are tried against each line determines which rule
matches.  Overlapping rules are frequently a requirement, allowing a more
specific rule to match some lines and a more general rule to match the
majority, e.g.\ separating \SMTP{} delivery to a filtering proxy from
\SMTP{} delivery to the rest of the world.  The algorithm provides a
facility for ordering overlapping rules: the priority field in each rule
(defaults to zero).  Rules are tried in order of priority, highest first,
overriding the hits field maintained by the parser.  Negative priorities
may also be useful for catchall rules.

@@

    \item Mail enters the system via \SMTP{} or local submission.

@@

Everything starts off with a mail entering the system, whether by local
submission via postdrop/sendmail, by \SMTP{}, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: pickup; flowchart:~2).

\SMTP{} is more complicated:

@@

    \item The client disconnects (action: disconnect; flowchart:~6).  If
        Postfix has rejected any \SMTP{} commands the data will be saved to
        the database; if not there won't be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

@@

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: track; flowchart:~11).\footnote{Relayed
        delivery is performed by the \SMTP{} client; local delivery means
        local to the server, i.e.\ an address the server is final
        destination for.} In this case the mail may already have been
        created (action: qmgr\_chooses\_mail; flowchart:~3; described
        above) and the unknown origin flag will be cleared; if not a new
        data structure will be created.  In both cases the new mail is
        marked as a child of the parent.  The log line in question is:

@@

    \item [REJECTION] Deal with postfix rejecting an \SMTP{} command from
        the remote client: log the rejection with a mail if there is a
        queueid in the log line, or with the connection if not.

@@

    \item Neither smtpd nor pickup have logged any messages associated with
        the mail, indicating it was generated internally by Postfix, not
        accepted via \SMTP{} or submitted locally by sendmail.

@@

Some mail clients appear to send the following sequence of commands during
the \SMTP{} session:

@@

The DATA phase of the \SMTP{} conversation is where the headers and body of
the mail are transferred.  Sometimes there is a timeout or the connection
is lost\footnote{For brevity's sake timeout will be used throughout this
section, but everything applies equally to lost connections.} during the
DATA phase; when this occurs Postfix will discard the mail and the parser
needs to discard the data associated with that mail.  It seems more
intuitive to save the data to the database, but if a timeout occurs there
won't yet have been any data gathered for the mail, so there is none
available to save; the timeout is saved with the connection data instead.

@@

            \item For mail sent: to addresses, \SMTP{} code, and delay.

@@

\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
size \& number of lines to execution time: statistics}{execution time vs
file size vs number lines factor table}

@@

The difference between logs with a mail loop and logs without can be seen
in table~\refwithpage{Rule ordering with and without a mail loop} showing
the increases for the different rule orderings and combinations of log
files:

\showtable{build/stats-normal-shuffle-reverse-include}{Rule ordering with
and without a mail loop}{Rule ordering with and without a mail loop}

@@

\showgraph{build/plot-cached-discarded-factor.ps}{Regex caching: percentage
execution time increase}{normal regex vs discarded regex factor}

@@

\showtable{build/stats-cached-discarded-include-for-graph}
{Regex caching/discarding with different groups of log files}
{Regex caching/discarding with different groups of log files}

@@

% This fools the syntax highlighting so we get a new fold here, but it
% doesn't create a new section.
\ifacronymfirstuse{SMTP}{
\section{Glossary}
}{}
% Redefine the command used to produce the glossary title, because the
% default command produces an unnumbered section whereas I want a numbered
% section.
\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
\printglossary{}
% Redefine the command a second time to produce acronyms instead of a
% glossary.
\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
\printacronym{}

% The contents of the glossary.
\glossary{name={ESMTP},description={
    ESMTP is Extended \SMTP{}, defined in \RFC{}~1869~\cite{RFC1869}.
}}
